---
title: "PB HLTH C240D/STAT C245D: Assignment #3"
author: "Rachael Phillips"
date: "November 28, 2017"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# *Maximum likelihood estimation of the ABO blood group allele frequencies using the EM algorithm*

The ABO *blood groups* were the first to be discovered and are important in assuring safe blood transfusions
(Cf. Landsteiner, 1930 Nobel Prize in Physiology and Medicine, \texttt{nobelprize.org/nobel\_prizes/medicine/laureates/1930})
As indicated in Table 1, the ABO blood groups are characterized by the presence or absence of *antigens* on the surface of red blood cells and *antibodies* in serum. The ABO locus has three *alleles*, A, B, and O, leading to $3^2$ = $9$ *phased genotypes*, $3 + 3 \times 2/2 = 6$ *unphased genotypes*, and four *phenotypes*, the blood groups **A**, **B**, **AB**, and **O**.

Let $\pi = (\pi_A,\pi_B,\pi_O)$ denote the ABO allele frequencies in a well-defined population of interest. Under the assumption of *Hardy-Weinberg equilibrium* (HWE), the maternal and paternal alleles are independent, i.e., genotype frequencies are products of allele frequencies. Let $Y = (Y_A,Y_B,Y_{AB},Y_O)$ denote the ABO phenotype counts for a random sample of $n$ individuals from the population of interest.

The objective of this assignment is to apply the EM algorithm to derive maximum likelihood estimates of the ABO allele frequencies for a dataset from the classical article of Clarke et al. (1959). Specifically, consider the following ABO phenotype counts for a sample of $n$ = 521 duodenal ulcer patients (Clarke et al., 1959, Table III): $Y_A$ = 186, $Y_B$ = 38, $Y_{AB}$ = 13, and $Y_O$ = 284. For simplicity, you may assume that the $n$ = 521 patients are a random sample from a well-defined population, with Hardy-Weinberg equilibrium at the ABO locus.

\vspace{.3in}

Table 1: ABO *blood groups*. Phenotypes, genotypes, and genotype frequencies under Hardy-Weinberg equilibrium.

| **Phenotype ABO blood group** | **Antigens** | **Antibodies** | **Unphased genotype** | **Unphased genotype frequency (HWE)** |
|:-------------------------:|:--------:|:-----------------------:|:-----------------:|:---------------------------------:|
| **A** | A | Anti-B | AA,AO | $\pi_A^2 + 2\pi_A\pi_O$ |
| **B** | B | Anti-A | BB, BO | $\pi_B^2 + 2\pi_B\pi_O$ |
| **AB** | A and B | Neither | AB | $2\pi_A\pi_B$ |
| **O** | Neither | Anti-A and Anti-B | OO | $\pi_O^2$ |

## Question 1. Conditional distribution of multinomial counts.

Suppose $X=(X_k:k=1, ..., K)$ is a random variable with Multinomial$(n, \pi = (\pi_k : k=1, ..., K))$ distribution, that is:

$$Pr(X=x) = \frac{n!}{\prod_{k=1}^K x_k!}\prod_{k=1}^K \pi_k^{x_k}  \ \ \ \ \ \ \ \ \ \ \ \text{(1)}$$

where $x = (x_k : k = 1, ..., K) \in \mathbb{N}^K$, with $\sum_k x_k = n$ and $\pi = (\pi_k : k=1, ... K) \in [0,1]^K$, with $\sum_k \pi_k = 1$.

Derive the conditional distribution of $X_k$ given $X_k + X_{k'}$, where $k,k' \in \{ 1, ... , K \}, k \neq k'$.

In particular, provide $E[X_k | X_k + X_{k'}]$, the conditional expected values of $X_k$ given $X_k + X_k'$.

\vspace{.1in}


## Question 2. Log-likelehood surface for trinomial probabilities

Propose and implement in R graphical displays for a *log-likelehood surface* when the parameter of interest is a vector of trinomial probabilities.

Display and comment on the log-likelehood surface for the ABO blood group phenotype counts in the Clarke et al. (1959) dataset.

**Hint.** Use barycentric coordinates.

\vspace{.1in}

```{r}

```

## Question 3. Derivation of EM algorithm.
Derive the *expectation-maximization algorithm* (EM) for *maximum likelihood estimation* (MLE) of the ABO allele frequencies $\pi = (\pi_A,\pi_B,\pi_O)$, based on ABO phenotype counts from a random sample of $n$ individuals from a well-defined population, with Hardy-Weinberg equilibrium at the ABO locus.

Specifically, define the observed incomplete and unobserved complete data structures, provide the incomplete and complete data log-likelihood functions, supply the main EM $Q$-function, and derive explicit solutions for the E- and M-steps.

\vspace{.1in}


## Question 4. Software implementation of EM algorithm.

Write an R function implementing the EM algorithm for maximum likelihood estimation of the ABO allele frequencies $\pi = (\pi_A,\pi_B,\pi_O)$.

Arguments to this function should include: the phenotype counts, starting values for the allele frequencies, stopping criteria; it should return candidate MLE for the allele frequencies and the corresponding value of the observed data log-likelihood.

\vspace{.1in}

```{r}

```

## Question 5. Application of EM algorithm.

Apply the EM algorithm to derive maximum likelihood estimates of the ABO allele frequencies $\pi = (\pi_A,\pi_B,\pi_O)$ for the Clarke et al. (1959) dataset.

Trace the progress of the EM algorithm by providing a table of candidate MLE for the allele frequencies and corresponding values of the observed data log-likelihood at each iteration.

Also provide graphical summaries of these results.

Comment on the EM algorithm’s performance in terms of sensitivity to starting values, convergence, and any other features you deem relevant.

Compare the results from your implementation of the EM algorithm to those from one of the R optimization functions (e.g., \texttt{optim}).

\vspace{.1in}


### Collaborators & References
